---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# survSAKK

<!-- badges: start -->
[![check-release](https://github.com/SAKK-Statistics/survSAKK/actions/workflows/check-release.yaml/badge.svg)](https://github.com/SAKK-Statistics/survSAKK/actions/workflows/check-release.yaml)
<!-- badges: end -->

The goal of `survSAKK` is to create an open source, user-friendly and flexible `R` package, which will incorporate various statistics and layout customization options to enhance the efficiency and adaptability of the Kaplan-Meier plot. 

## Installation

You can install the development version of survSAKK from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("SAKK-Statistics/survSAKK", build_vignettes = TRUE)
```

## Load package

```{r example}
library(survSAKK)
## basic example code
```

## Get the documentation of the funciton

```{r}
?survSAKK
```

## Access vignette
```{r}
vignette(survSAKK)
```



