---
title: "survSAKK::surv.plot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{survSAKK::surv.plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = TRUE
)
```

# Preparation

## Load libraries

```{r setup, warning=FALSE}
library(survSAKK)
library(survival)
```

## Load Data

```{r data}
veteran <- survival::veteran
veteran$time_yr <- veteran$time/365.25
veteran$time_mt <- veteran$time_yr*12

# Define the reference arm
veteran$trt <- factor(veteran$trt,
                      levels = c(1,2),
                      labels = c("Standard","Test"))

# Create survival object
veteran_fit_yr <- survfit(Surv(time_yr, status) ~ 1, data = veteran)
veteran_fit_mt <- survfit(Surv(time_mt, status) ~ 1, data = veteran)
veteran_trt_fit_yr <- survfit(Surv(time_yr, status) ~ trt, data = veteran)
veteran_trt_fit_mt <- survfit(Surv(time_mt, status) ~ trt, data = veteran)

```

# Examples of 'surv.plot' Function

## Base plot

```{r baseplot}
survSAKK::surv.plot(veteran_fit_yr)
```

## Adjustment of layout

```{r layout}
surv.plot(veteran_trt_fit_mt,
          col = c("darkblue", "black"),
          main = "Veterans' Administration Lung Cancer study", 
          xlab = "Time (months)", 
          legend.name = c("Standard","Test"))
```

## Drawing segment for specific timepoint/quantile

### Estimated survial probability at 0.5 quantile (median)

```{r quantil}
surv.plot(veteran_trt_fit_mt,
          col = c("darkblue", "black"),
          xlab = "Time (months)", 
          legend.name = c("Standard","Test"),
          segment.quantile = 0.5)
```

### Estimated survial probability at 6 month

```{r timepoint}
surv.plot(veteran_trt_fit_mt,
          col = c("darkblue", "black"),
          xlab = "Time (months)", 
          legend.name = c("Standard","Test"),
          segment.timepoint = c(6))
```

### Estimated survival probability at 3,6 and 9 month

```{r timepoints, eval=FALSE}
surv.plot(veteran_trt_fit_mt,
          segment.timepoint = c(3,6,9),
          segment.annotation = "none",
          segment.type = 2,
          segment.col = "darkblue")
```

## Inlcude statistics

### Cox proportional hazard 

```{r coxph}
surv.plot(veteran_trt_fit_mt,
          segment.quantile =  0.5,
          stat = "coxph",
          stat.position = "bottomleft")
```

## Modify risk table

### Undisplay risk table

```{r norisktable}
surv.plot(veteran_trt_fit_mt,
          segment.quantile =  0.5,
          stat = "coxph",
          stat.position = "bottomleft",
          risktable = FALSE)
```

### Adjustment of the risktable

```{r risktable}
surv.plot(veteran_trt_fit_mt,
          col = c("darkblue","black"),
          legend.name = c("Standard", "Test"),
          segment.quantile =  0.5,
          segment.annotation = c(17,0.9),
          stat = "coxph",
          stat.position = "bottomleft",
          risktable.col = c("darkblue","black"),
          risktable.title.font = 1,
          risktable.name.font = 2)
```


# Mix multiple 'surv.plot' plots

```{r plots}
par(mfrow=c(2,3))
# Plot 5
surv.plot(fit = veteran_fit_mt,
          risktable = FALSE,
          xlim = seq(0,34, by = 3),
          xlab = ("Time (month)"),
          segment.type = 3,
          segment.quantile = 0.5,
          segment.annotation.space = 0.06,
          segment.cex = 0.6,
          risktable.cex = 0.4,
          risktable.name.position = -4)
# Plot 6
surv.plot(fit = veteran_fit_yr,
          risktable = TRUE,
          xlim = seq(0,3, by = 0.5),
          xlab = c("Time (year)"),
          ylab = c("Survival"),
          main = "Kaplan-Meier Pots",
          legend.name = c("Veterans"),
          legend.cex = 0.6,
          segment.type = 3,
          segment.timepoint = 1.0,
          segment.annotation.space = 0.06,
          segment.annotation = c(1.58,0.75),
          segment.cex = 0.6,
          segment.col = "darkred",
          risktable.cex = 0.4,
          risktable.title.position = -0.5,
          risktable.name.position = -0.5)
# Plot 7
surv.plot(fit = veteran_trt_fit_mt,
          risktable = TRUE,
          legend.name = c("LT60","OV60"),
          legend.position = c(23.7,0.3),
          legend.cex = 0.6,
          xlim = seq(0,34, by = 6),
          xlab = ("Time (month)"),
          segment.timepoint = 6,
          segment.annotation.space = 0.06,
          segment.main = "Survival at 6mt [95%]",
          segment.annotation = c(18, 0.85),
          segment.cex = 0.6,
          segment.col = c("#666666","#a6761d"),
          stat = "coxph",
          stat.position = "bottomleft",
          stat.font = 2,
          stat.cex = 0.6,
          risktable.cex = 0.5)

# Plot 8
surv.plot(fit = veteran_trt_fit_mt,
          risktable = TRUE,
          grid =FALSE,
          col = c("pink","lightblue"),
          xlab = "Time (year)",
          xlim = seq(0,3),
          legend.title = "Treatment Regimens",
          legend.title.cex = 0.5,
          legend.name = c("LT60","OV60"),
          legend.cex = 0.6,
          stat = "coxmodel",
          stat.position = "bottomleft",
          stat.cex = 0.6,
          risktable.cex = 0.5
)

# Plot 9
surv.plot(fit = veteran_trt_fit_mt,
          risktable = TRUE,
          grid =TRUE,
          col = c("#5ab4ac","#d8b365"),
          xlab = "Time (year)",
          xlim = seq(0,3, by = 0.25),
          legend.title = "Treatment Regimens",
          legend.title.cex = 0.5,
          legend.name = c("LT60","OV60"),
          legend.cex = 0.6,
          stat = "coxph",
          stat.position = "bottomleft",
          stat.cex = 0.6,
          risktable.cex = 0.5,
          risktable.col = c("#5ab4ac","#d8b365"),
          risktable.name.font = 4
)

# Plot 10
surv.plot(fit = veteran_trt_fit_mt,
          risktable = FALSE,
          grid =TRUE,
          col = c("#fc8d59","#99d594"),
          xlab = "Time (year)",
          xlim = seq(0,3),
          legend.title = "Treatment Regimens",
          legend.title.cex = 0.5,
          legend.name = c("LT60","OV60"),
          legend.cex = 0.6,
          segment.type = 1,
          segment.quantile = 0.5,
          segment.annotation = "left",
          segment.annotation.space = 0.04,
          segment.cex = 0.5,
          segment.font = 2,
          stat = "logrank",
          stat.position = "bottomleft",
          stat.cex = 0.6,
          risktable.cex = 0.5,
          risktable.title = "No. at risk",
          risktable.title.font = 2)
          
par(mfrow=c(1,1))

```


### References
